---
title: "Gender Pay Gap: Individuatl Dataset"
output: html_document
---
DS1 Final Project   09.07.2020  Maki Terao
Part C: individual data set
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

library(tidyverse)
library(ggplot2)
library(corrplot)
library (readr)
library(plyr)
library(reldist)#for gini coefficients
```

Data is 2018 ACS PUMS Massachusetts State (This is for MA analysis, Script is same for WA and GA)
https://www2.census.gov/programs-surveys/acs/data/pums/2018/1-Year/
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

data=read_csv("psam_p25.csv", col_names=T)
#chose related variables. including child variable (I can only find OC and PAOC in this dataset)
ma=data[c("AGEP","HISP","NATIVITY","MAR","SEX","SCHL","SCHG","RELP","RAC1P","OCCP","WAOB", "OC","PAOC", "WAGP")]
```

View missing values
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
missing_data = ma%>% summarise_all(funs(sum(is.na(.))/n()))
missing_data = gather(missing_data, key = "variables", value = "percent_missing")
ggplot(missing_data, aes(x = reorder(variables, percent_missing), y = percent_missing*100)) +
  geom_bar(stat = "identity", fill = "dark blue", aes(color = I('white')), size = 0.3)+
  xlab('variables')+ylab('% missing')+ggtitle('% of Missing Variables')+
  coord_flip()+ 
  theme_bw()

```
Take a look SEX, 1 male, 2 female,  OC "child" , PAOC, "cild and their age" distribution of entire dataset. PAOC does not provide much information.  
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
ma$SEX=factor(ma$SEX)
table(ma$SEX)
ma$OC=factor(ma$OC)
table(ma$OC)
ma$PAOC=factor(ma$PAOC)
table(ma$PAOC)
```
Drop all observations with $0 WAGP (last 12 months wage) variable and take a look child, education and age variables for the group.
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
maworker=subset(ma, WAGP!=0)
summary(maworker$WAGP) 
table(maworker$OC)
summary(maworker$AGEP)
```
Create "ocpCat" occupation category (530 subcategories to 25 major cotegories)
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

maworker$OCCP=as.numeric(maworker$OCCP)

x=maworker$OCCP
maworker$ocpCat[x<=440]="MGR"
maworker$ocpCat[500<=x & x<=750]="BUS"
maworker$ocpCat[800<=x & x <= 960]="FIN"
maworker$ocpCat[1005<=x & x<=1240]="CMM"
maworker$ocpCat[1305<=x & x <= 1560]="ENG"
maworker$ocpCat[1600<=x & x<=1980]="SCI"
maworker$ocpCat[2001<=x & x <= 2060]="CMS"
maworker$ocpCat[2100<=x & x<=2180]="LGL"
maworker$ocpCat[2205<=x & x <= 2555]="EDU"
maworker$ocpCat[2600<=x & x<=2920]="ENT"
maworker$ocpCat[3000<=x & x <= 3550]="MED"
maworker$ocpCat[3601<=x & x <= 3655]="HLS"
maworker$ocpCat[3700<=x & x<=3960]="PRT"
maworker$ocpCat[4000<=x & x <= 4160]="EAT"
maworker$ocpCat[4200<=x & x<=4255]="CLN"
maworker$ocpCat[4330<=x & x <= 4655]="PRS"
maworker$ocpCat[4700<=x & x <= 4965]="SAL"
maworker$ocpCat[5000<=x & x<=5940]="OFF"
maworker$ocpCat[6005<=x & x <= 6130]="FFF"
maworker$ocpCat[6200<=x & x<=6765]="CON"
maworker$ocpCat[6800<=x & x <= 6950]="EXT"
maworker$ocpCat[7000<=x & x <= 7750]="RPR"
maworker$ocpCat[7800<=x & x<=8990]="PRD"
maworker$ocpCat[9005<=x & x <= 9760]="TRN"
maworker$ocpCat[9800<=x & x<=9830]="MIL"
maworker$ocpCat[9920==x]="UNE"

```
Drop annual wage value less than $23,920 (38,237->26,655)
Exclude part-time, less than full year workers.  Median increases $65,000
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
maworkerF=subset(maworker, WAGP >= 23920)
summary(maworkerF$WAGP) 
```
Take a look at SEX and OC after dropping less than 23,900 and median wage by sex.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
table(maworkerF$OC) #only 1 observation with child (unable to use this variable)

table(maworkerF$MAR)#1 Married~5 Never Married (see EDA below for description)

#by sex male , female
mwsex=ddply(maworkerF, .(SEX), summarize, medwage=median(WAGP))
mwsex
```

Calculate male/ female median wage, pay gap, and number of workers by ocpCat

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
mwlist=ddply(maworkerF, .(ocpCat), summarize, medwage=median(WAGP))
#median wage by ocpCat 
d=aggregate(maworkerF$WAGP, by=list(maworkerF$ocpCat,maworkerF$SEX), FUN=median)
colnames(d)=c("ocpCat", "SEX", "wage")
d=spread(d,SEX, wage)
mwlist=left_join(mwlist, d, by="ocpCat", dupliate=F)
names(mwlist)[3]="male"
names(mwlist)[4]="female"
mwlist$mapper=mwlist$female/mwlist$male
#num of workers by ocpCat
mwlist$ttlwk=table(maworkerF$ocpCat)
e=maworkerF %>% group_by(SEX, ocpCat) %>% tally()
e=spread(e,SEX,n)
mwlist=left_join(mwlist, e, by="ocpCat", duplicate=F)
names(mwlist)[7]="ttlml"
names(mwlist)[8]="ttlfl"

#write.csv(mwlist,"ocpCatma.csv", row.names=F)

```

Create age category "ageC"
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
summary(maworkerF$AGEP) #median 46, Mean 45.13 (higher than WA & GA)

x=maworkerF$AGEP
maworkerF$ageC[x<=24]="1"
maworkerF$ageC[25<=x & x<=34]="2"
maworkerF$ageC[35<=x & x <= 44]="3"
maworkerF$ageC[45<=x & x<=54]="4"
maworkerF$ageC[55<=x & x <= 64]="5"
maworkerF$ageC[65<=x]="6"
table(maworkerF$ageC)

```
Create race category "raceC" and "latino"
I will assume "HISP" variable's 02~24 are latino and within non-latino group, categorize by 
variable RAC1P.  RAC1P, I will categorize by 1.White, 2. Black 6. Asian, 3~5&7. Native, 8& 9.other

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
#table(maworkerF$RAC1P) #no hispanic category
#table(maworkerF$HISP)  #01: not hispanic 02~24: hispanic

x=maworkerF$HISP
maworkerF$latino[x=="01"]="0" #not-hispanic/ latino  
maworkerF$latino[x!="01"]="1" #hispanic 
#table(maworkerF$latino)# 0=24,928, 1=1727

x=maworkerF$RAC1P
y=maworkerF$latino
maworkerF$raceC[y!=1 & x==1]="1" #white 
maworkerF$raceC[y!=1 & x==2]="2" #black
maworkerF$raceC[y!=1 & x==6 ]="3" #asian
maworkerF$raceC[y!=1 & (x==3|x==4|x==5|x==7)]="5"#native
maworkerF$raceC[y!=1 & (x==8|x==9)]="6"#other
maworkerF$raceC[y==1]="4"#hispanic or latino
table(maworkerF$raceC)

```
Create education attainment category "edC" see EDA for category description. 
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
x=maworkerF$SCHL
maworkerF$edC[x<=15]="1"
maworkerF$edC[16<=x & x<=17]="2"
maworkerF$edC[18<=x & x <= 20]="3"
maworkerF$edC[x==21]="4"
maworkerF$edC[x==22]="5"
maworkerF$edC[x==23]="6"
maworkerF$edC[x==24]="7"
maworkerF$edC=as.numeric(maworkerF$edC)
hist(maworkerF$edC)
```
Take a look at variance, etc.
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
var(log(maworkerF$WAGP))
sd(maworkerF$WAGP)
gini(maworkerF$WAGP)

#write.csv(maworkerF,"maworkerF.csv",row.names=F)
#maworkerF=read.csv("maworkerF.csv")
```
EDA
Total population by SEX: (remove package "plyr") 

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
detach("package:plyr", unload = TRUE)

par(mfrow=c(1,4))
#all dataset
data %>% 
  group_by(SEX) %>% 
  summarise(Count = n())%>% 
  mutate(percent = prop.table(Count)*100)%>%
  ggplot(aes(x=SEX,y=percent), fill = SEX)+
  geom_col(fill = c("grey", "light blue"))+
  geom_text(aes(label = sprintf("%.1f%%", percent)), hjust = 0.2, vjust = 2, size = 5)+ 
  theme_bw()+  
  xlab("SEX 1:male  2:female") + ylab("Percent") + ggtitle("MA Population by SEX")
#workers only
maworkerF %>% 
  group_by(SEX) %>% 
  summarise(Count = n())%>% 
  mutate(percent = prop.table(Count)*100)%>%
  ggplot(aes(x=SEX, y=percent), fill = SEX)+
  geom_col(fill = c("grey", "light blue"))+
  geom_text(aes(label = sprintf("%.1f%%", percent)), hjust = 0.2, vjust = 2, size = 5)+ 
  theme_bw()+  
  xlab("SEX 1:male  2:female") + ylab("Percent") + ggtitle("MA Worker (full) by SEX")

#tech (CMM) only subset
maCMM=subset(maworkerF, ocpCat=="CMM") #1562 ttl
maCMM %>% 
  group_by(SEX) %>% 
  summarise(Count = n())%>% 
  mutate(percent = prop.table(Count)*100)%>%
  ggplot(aes(x=SEX, y=percent), fill = SEX)+
  geom_col(fill = c("grey", "light blue"))+
  geom_text(aes(label = sprintf("%.1f%%", percent)), hjust = 0.2, vjust = 2, size = 5)+ theme_bw()+  
  xlab("SEX 1:male  2:female") + ylab("Percent") + ggtitle("Tech Worker by SEX")

#high wage only
maHW=subset(maworkerF, WAGP>=150000)
maHW %>% 
  group_by(SEX) %>% 
  summarise(Count = n())%>% 
  mutate(percent = prop.table(Count)*100)%>%
  ggplot(aes(x=SEX, y=percent), fill = SEX)+
  geom_col(fill = c("grey", "light blue"))+
  geom_text(aes(label = sprintf("%.1f%%", percent)), hjust = 0.2, vjust = 2, size = 5)+ theme_bw()+  
  xlab("SEX 1:male  2:female") + ylab("Percent") + ggtitle("150K+ Workers by SEX")


```
Median Wage by SEX
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
mwsex %>% ggplot(aes(x=SEX, y=medwage))+geom_col(fill=c('grey','light blue'))+
  geom_text(aes(label=paste0("$", medwage)), hjust = 0.2, vjust = 2, size = 5)+
   theme_bw()+  
  xlab("SEX 1:male  2:female") + ylab("") + ggtitle("Median Wage by SEX")
```
Number of workers & median wage by occupational category
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
par(mfrow=c(1,2))
#num of workders by Occupational Category 
maworkerF %>% group_by(ocpCat) %>% summarise(ocp=n()/1000)  %>% 
  ggplot(aes(x=reorder(ocpCat, ocp),y=ocp))+geom_bar(stat='identity', fill='dark blue')+
  coord_flip()+
  labs(x='Occupational Category', y='Number of Workers (thousands)')

#median wage by occp category
mwlist %>% ggplot(aes(x=reorder(ocpCat, medwage), y=medwage/1000))+
  geom_bar(stat='identity', fill='dark blue')+coord_flip()+
  labs(x='Occupational Category', y='Median Wage (100K)')

```
Number of workers by education attainment
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
par(mfrow=c(1,2))
ed_class=c('1'='Less than a high school diploma', '2'='High school graduate or GED', '3'='Some college or associate degree', 
           '4'="Bachelor's degree", '5'="Master's degree", '6'="Professional degree", '7'="PhD")
#all workers
maworkerF %>% group_by(edC) %>% summarize(n=n()) %>% 
  ggplot(aes(x=reorder(ed_class[edC], -edC), y=n/1000))+geom_bar(stat='identity', fill='dark blue')+
  labs(x='', y='Number of Workers (thousands)', title='All Occupations' )+
  coord_flip()

#tech only
maCMM %>% group_by(edC) %>% summarize(n=n()) %>% 
  ggplot(aes(x=reorder(ed_class[edC], -edC), y=n))+geom_bar(stat='identity', fill='dark blue')+
  labs(x='', y='Number of Workers', title='Tech Only' )+
  coord_flip()

```
Number of workers by age group
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

age_class=c('1'='16 to 24', '2'='25 to 34', '3'='35 to 44', '4'='45 to 54', '5'='55 to 64', '6'='65 and older')
maworkerF %>% group_by(ageC) %>% summarize(n=n()) %>% 
  ggplot(aes(x=age_class[ageC], y=n/1000))+geom_bar(stat='identity', fill='dark blue')+
  labs(x='', y='Number of Workers (thousands)', title='All Occupations' )
maCMM %>% group_by(ageC) %>% summarize(n=n()) %>% 
  ggplot(aes(x=age_class[ageC], y=n))+geom_bar(stat='identity', fill='dark blue')+
  labs(x='', y='Number of Workers', title='Tech Only' )

```
Change ageC, SEX, raceC(latino), MAR, Nativity variables to factor 
```{r}
mdata=maworkerF[c('ageC','SEX','edC','raceC','MAR','NATIVITY','ocpCat','OCCP','WAGP')]
mdata[,c(1:8)]=lapply(mdata[c(1:8)], as.factor)
#all with WAGP
fit1=lm(WAGP~.-OCCP, data=mdata) 
summary(fit1)
```
All workers with log WAGP (better R^2)

```{r}
fit2=lm(log(WAGP)~.-OCCP, data=mdata)
summary(fit2)
```

All workers all 530 OCCP facotrs & log WAGP +OCCP (MAR and NATIVITY did not show so I changed variables order)
```{r}
fit3=lm(log(WAGP)~.-ocpCat, data=mdata)#adj R2=0.4488 SEX2=-0.1914
summary(fit3)
 
```
Exclude high wage group 
n=23,996 (90.02% of total maworkerF) R2 0.3435 SEX2=-0.1315

```{r}
mdataP=subset(mdata, WAGP <=150000)
fit5=lm(log(WAGP)~.-OCCP, data=mdataP)
summary(fit5)
table(mdataP$SEX)
#                 1     2 
#<=150,000 group  12228 11768 
#>150,000  group  1995  664 

```
Single only 
n=7399 (m 3925, f3463)

```{r}
mdataS=subset(mdata, MAR=="5")
table(mdataS$SEX)

mdataS=mdataS[,-5]
fitS=lm(log(WAGP)~.-OCCP,data=mdataS)
summary(fitS)#R2=0.3261  SEX2=-0.08406
summary(mdataS$WAGP)#median wage goes down $50,000
summary(mdataS$ageC)#younger distribution 
```
Compare single women and all male group
n=17,686

```{r}
mdataM=subset(mdata, (SEX=="1"|(SEX=="2"& MAR=="5")))
table(mdataM$SEX)
mdataM=mdataM[,-5]

#with all OCCP R2=0.46, SEX2=-0.1877
fit6=lm(log(WAGP)~.-ocpCat, data=mdataM)
summary(fit6)
```


```{r}
#with ocpCat only R2=0.4079, SEX2=-0.2276
fit7=lm(log(WAGP)~.-OCCP, data=mdataM)
summary(fit7)

```
To see impact of marital status for both sex group
```{r}
#female only n=12,432
mdataF=subset(mdata, SEX=="2") 
mdataF=mdataF[,-2]
fit8=lm(log(WAGP)~.-OCCP, data=mdataF)
summary(fit8)#R2=0.3547  MAR5=-0.0737
```
```{r}
#male only n=14,223
mdataE=subset(mdata, SEX=="1") 
mdataE=mdataE[,-2]
fit10=lm(log(WAGP)~.-OCCP, data=mdataE)
summary(fit10) #R2=0.3982  MAR5=-0.2469

```
Create coefficient data table
```{r}
stat01 = coef(summary(fit2))[, c("Estimate","t value")] %>% as.data.frame()
stat01=rownames_to_column(stat01, var="variables")
stat02=coef(summary(fit3))[, c("Estimate","t value")] %>% as.data.frame()
stat02=rownames_to_column(stat02, var="variables")
coefdata=left_join(stat01, stat02, by="variables")
colnames(coefdata)=c("p", "Estimate", "t value", "Estimate OCCP", "t value OCCP")

#write.csv(coefdata, "coefdata1.csv", row.names = F)
```
For all male and single female result AND single male and single female result 

```{r}
stat07 = coef(summary(fit7))[, c("Estimate","t value")] %>% as.data.frame()
stat07=rownames_to_column(stat07, var="variables")
stat06=coef(summary(fit6))[, c("Estimate","t value")] %>% as.data.frame()
stat06=rownames_to_column(stat06, var="variables")
statS=coef(summary(fitS))[, c("Estimate","t value")] %>% as.data.frame()
statS=rownames_to_column(statS, var="variables")

coefdata2=left_join(stat07, stat06, by="variables")
coefdata2=left_join(coefdata2, statS, by="variables")
colnames(coefdata2)=c("p", "Estimate", "t value", "Estimate OCCP", "t value OCCP","Estimate S", "t value S" )
#write.csv(coefdata2, "coefdata2.csv", row.names = F)

```


```
I run this code for project but it is getting really long so I did not show result. 
Result shown in Power Point. 

#tech job only n=1562
tecdata=maCMM[c('ageC','SEX','edC','raceC','OCCP','MAR','NATIVITY','WAGP')]
table(tecdata$SEX)
tecdata[,c(1:7)]=lapply(tecdata[c(1:7)], as.factor)
fitA=lm(log(WAGP)~., data=tecdata) #adj R2=0.3632
summary(fitA)

#other occupation OFF
maOFF=subset(maworkerF, ocpCat=="OFF") #2765 ttl
table(maOFF$SEX)#male 787, female 1978
offdata=maOFF[c('ageC','SEX','edC','raceC','OCCP','MAR','NATIVITY','WAGP')]
offdata[,c(1:7)]=lapply(offdata[c(1:7)], as.factor)
str(offdata)
fitC=lm(log(WAGP)~., data=offdata)
summary(fitC) #R2=0.2468

#managers
maMGR=subset(maworkerF, ocpCat=="MGR") #6341 ttl
table(maMGR$SEX)#male 3495, female 2846
mgrdata=maMGR[c('ageC','SEX','edC','raceC','OCCP','MAR','NATIVITY','WAGP')]
mgrdata[,c(1:7)]=lapply(mgrdata[c(1:7)], as.factor)
str(mgrdata)
fitF=lm(log(WAGP)~., data=mgrdata)
summary(fitF) #R2=0.3075

#sales
maSAL=subset(maworkerF, ocpCat=="SAL") #1994 ttl
table(maSAL$SEX)#male 1248, female 746
saldata=maSAL[c('ageC','SEX','edC','raceC','OCCP','MAR','NATIVITY','WAGP')]
saldata[,c(1:7)]=lapply(saldata[c(1:7)], as.factor)
summary(saldata) 
fitH=lm(log(WAGP)~., data=saldata)
summary(fitH) #R2=0.3504


#coefficient data table by ocp categories
statA = coef(summary(fitA))[, c("Estimate","t value")] %>% as.data.frame()
statA=rownames_to_column(statA, var="variables")
statC=coef(summary(fitC))[, c("Estimate","t value")] %>% as.data.frame()
statC=rownames_to_column(statC, var="variables")
statF = coef(summary(fitF))[, c("Estimate","t value")] %>% as.data.frame()
statF=rownames_to_column(statF, var="variables")
statH=coef(summary(fitH))[, c("Estimate","t value")] %>% as.data.frame()
statH=rownames_to_column(statH, var="variables")

coefdata3=list(statA,statC,statF,statH) %>% reduce(merge, by="variables", all.x=F, no.dups=T)
colnames(coefdata3)=c("p","CMM","CMM t-value", "OFF", "OFF t-value", "MGR", "MGR t-value", "SAL", "SAL t-value")

write.csv(coefdata2, "coefdatama3.csv", row.names=F)


```


